### 2018-03-16 paper/4-build.tex
Work on S4

The focus of this paper is on a variety of implementations of \hs{Build}, given
a \emph{client-supplied} implementation of \hs{Task}. That is, we are going to take
\hs{Task} as given from now on, and explore variants of \hs{Build}: first
abstractly (in this section) and then concretely in~\S\ref{sec-implementations}.

\todo{SLPJ}{What does this para add?}
Let \hs{result}~\hs{=}~\hs{build}~\hs{task}~\hs{key}~\hs{store}. The build
system \hs{build} is responsible for ensuring that \hs{key} and its dependencies
are up to date in \hs{result} starting from \hs{store} (\S\ref{sec-build-correctness}). As part of that work,
\hs{build} \emph{may} store auxiliary information \hs{info} of type \hs{i} in
the \hs{store}, which will be provided to future calls of \hs{build}.

\todo{SLPJ}{What does this para add?}
All the build systems we define are polymorphic in both \hs{k} and \hs{v}
(perhaps adding constraints to them), and have specific instantiations of \hs{c}
and \hs{i}. As an example, recall that the \hs{busy} build system defined
in~\S\ref{sec-general-build}, has \hs{c}~\hs{=}~\hs{Monad}, i.e. it supports
monadic tasks, and \hs{i = ()}, i.e. it does not store any build information.

According to the definition of minimality~\ref{def-minimal}, a minimal build
system must \textbf{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement is to \textbf{build all keys in
the order that respects their dependencies}.

We have emphasised two different phrases above, and tackle each aspect separately.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) is to simply build dependencies when they are
requested. By combining that with a transient log of which keys have already
been built, you can obtain a minimal build system.

@@

The idea of a dirty bit is to have one piece of persistent
information per key, saying whether the key is
\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
next build starts anything that changed between the two states is marked
dirty.
% ; and by marking additional things dirty/clean the build system can track
% what needs to rebuild.
When reaching a key, if it and all its dependencies are
clean, the key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell it (and its dependencies) dirty
if the user modifies it.
When rebuilding,
if a cell only depends on clean cells it is skipped, otherwise it is rebuilt and
marked dirty itself.
\todo{SLPJ}{What does ``marked dirty itself'' mean?}

@@

With a dirty bit it is possible to achieve minimality. To achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could model this
approach, but does not. In contrast, \Make cannot model early cutoff -- to do so
it would have to mark the node clean (so it didn't rebuild in the next run) and
at the same time not mark the things it depends on dirty -- an impossible task
with only the ability to update to the latest modification time.

@@

An alternative way to determine if a key is dirty is to record what state the
values/hashes of dependencies were used at last time, and if something has
changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
which we can use to \textit{verify} existing values. We can describe a trace as:

\todo{AM/NM}{Change from \hs{[Trace k v]} to \hs{Map k (Trave k v)} or something
similar, so that it's obvious we have one trace per key .}

\begin{minted}{haskell}
data Trace k v = Trace
    { key          :: k
    , dependencies :: [(k, Hash)]
    , result       :: Hash }
\end{minted}

We assume that \hs{Hash} is a small constant size, constructed from hashing the
underlying \hs{v} rather than storing it directly. Checking a trace requires
ensuring all the dependencies are up to date (using whatever ordering strategy
as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
same as the current value and that the result is the same. After computing a
fresh value we add its \hs{Trace} to the list. Therefore the information stored
by a build system that verifies traces is \hs{[Trace k v]}.

@@

A verifying trace allows us to mark a key dirty and rebuild it. Extending that information we can store a \textit{constructive} trace which is the trace plus the actual result. Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality. We can represent that as:

\todo{AM/NM}{Traces -- why the indirection (first get the hash, then look it up)?
Can we drop it? Can we switch to a more efficient/explicit data structure than a list?}

@@

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \textit{without} recomputing it.

@@

\todo{AM}{AM: I would also like to see actual \hs{verify} and \hs{construct} functions.

NM: It actually gets a bit more confusing than that - since you end up needing to pass functions to get yourself, your dependencies etc. Unclear there is a win. The functions would occur in the next section anyway.

AM: Now, after we've switched to \hs{Store i k v} it might work better?}

\subsection{Summary} \label{sec-design-space}

\todo{AM/NM/SLPJ}{Add a decription for the table.}

>>>>

The focus of this paper is on a variety of implementations of
\hs{Build}~\hs{c}~\hs{i}~\hs{k}~\hs{v}, given
a \emph{client-supplied} implementation of \hs{Task}~\hs{c}~\hs{k}~\hs{v}. That
is, we are going to take \hs{Task} as given from now on, and explore variants of
\hs{Build}: first abstractly (in this section) and then concretely
in~\S\ref{sec-implementations}.

As per the definition of minimality~\ref{def-minimal}, a minimal build
system must \underline{rebuild only out-of-date keys} and at most once. The only
way to achieve the ``at most once'' requirement while producing a correct build
result (\S\ref{sec-build-correctness}) is to \underline{build all keys in the
order that respects their dependencies}.

\vspace{1mm}
We have underlined two different phrases above, and tackle each aspect separately.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}), is to simply build dependencies when they are
requested. By combining that with a transient log of which keys have already
been built, you can obtain a minimal build system.

@@

The idea of a dirty bit is to have one piece of persistent
information per key, saying whether the key is
\emph{dirty} or \emph{clean}. After a build, all bits are set to clean. When the
next build starts, anything that changed between the two states is marked
dirty.
% ; and by marking additional things dirty/clean the build system can track
% what needs to rebuild.
When reaching a key, if it and all its transitive dependencies are clean, the
key does not need recomputing.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
When rebuilding, if a cell only depends on clean cells it is skipped, otherwise
it is rebuilt and is marked dirty so that the cells that depend on it are
subsequently rebuilt too.

@@

With a dirty bit it is possible to achieve minimality. To achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff -- to
do so it would have to mark the node clean (so it would not rebuild in the next
run) and at the same time not mark the things it depends on dirty -- an
impossible task with only the ability to update to the latest modification time.

\todo{NM}{I stopped here, as the traces seem not finalised yet.}

@@

An alternative way to determine if a key is dirty is to record what state the
values/hashes of dependencies were used at last time, and if something has
changed, the key is dirty and must be recomputed -- in essence a \emph{trace}
which we can use to \emph{verify} existing values. We can describe a trace as:

\begin{minted}{haskell}
data Trace k v = Trace
    { key          :: k
    , dependencies :: [(k, Hash v)]
    , result       :: Hash v }
\end{minted}

We assume that \hs{Hash v} is a small constant size, constructed from hashing the
underlying \hs{v} rather than storing it directly. Checking a trace requires
ensuring all the dependencies are up to date (using whatever ordering strategy
as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
same as the current value and that the result is the same. After computing a
fresh value we add its \hs{Trace k v} to the list. Therefore the information stored
by a build system that verifies traces is \hs{[Trace k v]}.

\todo{NM}{We need to emphasise that lists are convenient for expressing the semantics,
but actual implementation uses a data structure optimised for a specific build system.}

@@

A verifying trace allows us to mark a key dirty and rebuild it. Extending that information we can store a \emph{constructive} trace which is the trace plus the actual result. Once we are storing the complete result it makes sense to record many constructive traces per key, and to share them with other users, providing cloud-build functionality. We can represent that as:

\todo{NM}{Clarify why the indirection (first get the hash, then look it up)?}

@@

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it.

@@

\subsection{Summary}\label{sec-design-space}

\todo{NM}{Check the table and add a decription.}
